---
title: "Regular expressions for R users"
author: "Andrew Gard"
format: html
embed-resources: true
---

Regular expressions let you search and manipulate strings based on patterns rather than exact matches. 

Throughout this vid, we'll make use of the `stringr` functions `str_subset`, `str_detect`, and `str_extract`, all of which are covered in a separate video. We'll also use the included character vectors `sentences` and `fruit`. 

```{r}
library(stringr)

sentences <- sentences[1:10]
sentences

fruit
str_subset(fruit, "berry")
```

## Matching special characters

The following characters have special meanings and can't be searched literally.

. \ | ( ) [ { ^ $ * + ?

For instance, a dot matches with *any* character, which explains the following output.

```{r}
str_extract(sentences, ".")
```

In order to match these metacharacters, precede them with two slashes, like this:

```{r}
str_extract(sentences, "\\.")
```

## Position and relative position

To specify that a substring comes at the beginning of a string, use `^`. To specify that it comes at the end, use `$`.

```{r}
str_subset(fruit, "^re")
str_subset(fruit, "er$") # note order
```

## Character sets

Use brackets to allow multiple options for the character to be matched.

```{r}
str_subset(sentences, "[oe]n")
str_extract(sentences, "[oe]n")
```

Ranges of letters and numbers are allowed.

```{r}
str_subset(sentences, "[a-j]o")
str_extract(sentences, "[a-j]o")
```

Inside of brackets, "^" represents negation.

```{r}
str_extract(sentences, "[^a-j]o")
```

## Character classes

It's also possible to specify classes of characters. "\d" matches any digit and "\w" matches and word character (a-z, A-Z, 0-9, and underscore), while "\D" and "\W" will match anything *except* digits and word characters, respectively. Many other classes are available. Enter `?regex` for a full list. 

## The role of `\` in Quarto

Quarto doesn't render expressions like "\d"  in the last section properly because it interprets the slash as an escape character. To actually render this string, use "\\d" instead. 

Similarly, to render a slash, use \\.

## Matching whitespace

Use `\\s` for space, `\\t` for tab, and `\\n` for a new line. 

```{r}
str_subset(sentences, "e\\sb")
```

## Repetition

Use curly braces to indicate how many times in a row a character must appear.

```{r}
str_subset(sentences, "e{2}")
```

Use a "+" to indicate at least one occurrence, possibly more. 

```{r}
str_subset(sentences, "she+t")
```

To indicate a character is optional, use "?". For instance, the next command detects both "the" and "often".

```{r}
str_extract(sentences, "th?e")
```

Finally, "*" indicates an optional character that may occur more than once.

```{r}
str_extract(sentences, "she*")
```

Notice that these symbols come *after* the characters they refer to, unlike "\" which comes before.

## Combining regex with `|`

Use the pipe to indicate an "or" statement. Expressions of either side are grouped by default.

```{r}
str_subset(sentences, "chicken|hog")
```

You can still group using parentheses, though. 

```{r}
str_subset(sentences, "(chicken|hog)s")
```

## A small disclaimer

When writing this document, I've prioritized the appearance of the .qmd file, not the rendered .html.



