---
title: "Dates and times"
author: Andrew Gard
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
theme_set(theme_minimal())

optical <- read_csv("optical.csv")
```

## Dates

Dates are best stored using the `Date` class, which `read_csv()` generally assigns when a set is imported. Consider `optical$dob`, for instance.

```{r}
glimpse(optical)
```

Internally, dates are stored as integers representing the number of days since 1 January 1970, as you can check using `as.numeric()` or `unclass()`. 

```{r}
unclass(optical$dob) |> head()
```

Dates can be subtracted to yield time differences, typically measured in days. 

```{r}
diff(optical$dob) 
```

## Parsing dates

Often, imported data sets include strings that look like dates but aren't automatically stored that way. The `as.Date()` function isn't a particularly user-friendly way forward, so we recommend the `tidyverse` function `lubridate::mdy()` and its friends `ymd()`, `ydm()`, `myd()`, `dmy()`, and `dym()`. These functions do an excellent job of parsing a wide variety of dating conventions, so long as the order of the components is consistent.

```{r}
text_dates <- c("Aug 3, 2026",
                "August 3 2026",
                "8/3/2026",
                "08-03-2026",
                "080326")

date_dates <- mdy(text_dates) 
date_dates

class(date_dates)
```

Dates are displayed in year-month-day format by default, which is ideal for sorting. If you need a different printing convention, use the `format()` function. 

Dates can be built manually using `make_date()`, which is particularly useful when a dataset includes separate columns for `year`, `month`, and `day`. 

```{r}
make_date(year = 2026, 
          month = 8,
          day = 3)
```

## Extracting components

Functions to extract date components all have logical names.

```{r}
weekdays(optical$dob) # base R

year(optical$dob) # lubridate
month(optical$dob)
day(optical$dob) 
```

## Date-times

Moments in time are identified using the `dttm` class. In the `optical` dataset, `appt_date` has this class. 

```{r}
glimpse(optical)
```

Date-times are relative to location. To be truly unambiguous, a time zone must be included. In general, R will default to UTC (also called Greenwich mean time), though there are cases where your machine's local time zone will be used instead. 

The `tz()` function is used to extract or assign time zone information. 

```{r}
tz(optical$appt_date)
tz(optical$appt_date) <- "America/Chicago"
```

Alternatively, you might want to convert times rather than just overwrite the location. Use `with_tz()` for this. 

```{r}
optical$appt_date |> head()
with_tz(optical$appt_date, "UTC") |> head()
```

Use `OlsonNames()` to see a list of approximately 600 available zones, and `Sys.timezone()` to see which one your system is currently using. To see your machine's local time, use `Sys.time()`.

## Visualization

`ggplot` formats date-times logically in all plots. 

```{r}
#| message: false

ggplot(optical, aes(x = appt_date)) + 
  geom_histogram()
```

If you need to change the format of the labels, use `scale_x_date`.


